KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-05-13 paper_overleaf/main.tex

\noindent Inferring familial relationships is important in population genetics. Knowing genetic kinship of ancient individuals leads to insights into the culture and social hierarchy of ancient populations, and is relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA is difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, or contamination from various sources.

Here, we present KIN, a Hidden-Markov-Model-based approach to identify identity-by-descent fragments and to estimate degree of relatedness. KIN can accurately determine up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage. KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses them to improve classification accuracy.

We show in simulations that KIN performs at near perfect classification accuracy at inferring first-degree relationships, and at 90\% accuracy for second degree even at 0.05x coverage.

We illustrate the utility of KIN on two applications, one on a Neandertal group from Siberia, Russia, and one in a group of Bronze-Age people from the Lech Valley, Germany. 
We find that in both cases, KIN matches the findings of existing approaches and detects several additional related individuals.

Thus, KIN is a useful tool for analyzing relatedness for large ancient-DNA datasets with low-coverage sequencing data that is applicable to any diploid species, without the requirement for a reference panel.
\end{abstract}

@@

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), i.e. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent will share exactly one set of chromosomes IBD with their offspring, while subsequent recombination means that a grandparent will, on average share a quarter of their genome with a grand-child. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}). These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

@@

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH). In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences ($p$) along the chromosome in both the cases $p$ can take values of $p_0$, $p_1$ or $p_2$ which are the expected proportion of differences in IBD states 0,1 and 2 respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

\section{New Approaches}\label{new_approaches}
To infer the case of relatedness, KIN fits one HMM for each pair of individuals and for each possible relatedness case. The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account. The best-fit model is then assigned as the inferred relatedness. If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$). Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \texttt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Model description}\label{method_overview} 
The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes. For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical). We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood. We also return the most likely locations of IBD tracts (using standard Viterbi algorithm), and the state posterior probabilities $\BZ$.

@@

To test the performance of KIN, and compare it to a commonly used method, READ \cite{kuhn_estimating_2018}, we simulated an ancient population (sampled from 2500 generations ago) with msprime \cite{kelleher_efficient_2016}, and artificially mated some of the individuals to form a pedigree (figure \ref{figS10:pedigree}). We downsampled this data to low coverages (4$x$ - 0.03$x$), and introduced ascertainment bias and contamination from modern humans (see section \ref{method}) to understand how these factors affect the performance of KIN and READ. Finally we apply KIN to published data and show that it can give reliable information about genetic kinship.

\paragraph{Evaluating the ROH detection}
We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}). We observe that as coverage decreases, the raw heterozygosity estimates become noisier. In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all  windows (except for the case of 0.1x coverage where we miss-classify 2 windows out of 220). We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x (Table \ref{tab:Table 2}).

@@

\paragraph{Evaluating IBD prediction}
We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. \ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd}$ Degree, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03x coverage depth.

@@

\paragraph{Evaluation of KIN on simulations and comparison to READ}
We next compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation) under four different simulation scenarios (Fig. \ref{fig3:Comparison_READ_KIN}). We show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data. One exception is  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We also find that READ is strongly impacted by contamination ($<3\%$), but that the correction implemented in KIN is sufficient to remove this bias. For simulations with ROH, we find the overall classification accuracy is only slightly reduced. Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN (for details, see table S4).

@@

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient data, we applied it to a Neandertal data set  from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}. This dataset contains genetic data from  14 skeletal remains that likely belong to contemporary Neandertals who occupied the cave between 63 ± 4 and 48 ± 3 kya \cite{laurits_skov_genetic_nodate}. DNA extracted from each of these remains were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}, and sequenced  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}. We further restricted these data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} to make the data for our relatedness analyses match that used by the authors \cite{laurits_skov_genetic_nodate}. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). Further, we identify Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.

@@

\subsubsection{Application to ancient modern humans}

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin. We found that in most cases the three methods agree, as only $<5\%$ of pairs (311 out of 6903) show a different relatedness for KIN and READ (table S2). However, most of these are at very low confidence for both the methods ($LL<1$ for KIN and $sd<1$ for READ). There are 28 pairs where KIN has $\Delta LL>1$. When we investigated these comparisons in more detail we found that in many cases, the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig. S\ref{figS8:eg1}). Only 27 pairs in this subset of 311 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs). Among these 27 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 where READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree, and pair AITI72-AITI77A where READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).
We further found 43 pairs for which READ estimates First-Degree, and compared the estimates of lcMLkin and KIN for these cases (Table S3). There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN. Among the 19 overlapping pairs, we found four pairs where KIN and lcMLkin differ, and again the heterozygosities along the genome match the inference of KIN (fig. S\ref{figS9:eg2}).

>>>>

\noindent Inferring genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA is difficult due to low coverage, ascertainment bias, or contamination from various sources.

Here, we present KIN, a Hidden-Markov-Model-based approach to estimate the relatedness of a pair of individuals and to identify the identical-by-descent segments they share. KIN can accurately classify up to 3rd-degree relatives and differentiate between sibling and parent-child relationships using ancient DNA sequence data with as little as 0.05x coverage. KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses them to improve classification accuracy.

We show with simulations that KIN infers first-degree relationships with nearly perfect accuracy, and second degree relationships with 90\% accuracy, even at 0.05x coverage.

To illustrate its utility, we apply KIN to a Neandertal group from Siberia, Russia, and a group of Bronze-Age people from the Lech Valley, Germany. We find that in both cases, KIN matches the findings of existing approaches and detects several additional related individuals.

KIN is a useful tool for analyzing relatedness for ancient DNA datasets of any size, even with low coverage, and is applicable to any diploid species, without the requirement for a reference panel.
\end{abstract}

@@

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), i.e. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent will share exactly one set of chromosomes IBD with their offspring, while subsequent recombination means that a grandparent will, on average share a quarter of their genome with a grand-child. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: the individuals share zero, one or two chromosomes IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}). These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

@@

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH). In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences ($p$) along the chromosome. In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

\section{Algorithm}\label{new_approaches}
To infer relatedness, KIN fits one HMM for each pair of individuals and for each possible relatedness. The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account. The best-fit model is then assigned as the inferred relatedness. If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$). Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \texttt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Model description}\label{method_overview} 
The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes. For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{rd}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical). We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood. We also return the most likely locations of IBD tracts (using standard Viterbi algorithm), and the state posterior probabilities $\BZ$.

@@

\subsection{Evaluation with simulations}
We first tested the performance of KIN with simulated pedigrees. We performed coalescent simulations to generate 7 unrelated diploid genomes and artificially mated them to form pedigrees of 17 individuals with relationships up to $5^{th}$ degree (figure \ref{figS10:pedigree}). To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see \ref{method}). To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that about $17\%$ of the genome is ROH. We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see \ref{method}). We created 60 pedigrees for each combination of average coverage and scenarios of presence/absence of ROH, contamination, and ascertainment bias, totalling 2,880 pedigrees.

\paragraph{ROH detection}
We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}). We computed the proportion of windows that are in ROH and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows. Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}). With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.004$). The largest bias observed ($9\%$) was for the datasets that included contamination. Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}). Surprisingly, the bias was lower for lower coverages, perhaps because there is only one possible ROH state per window to model a continuous variable, viz. proportion of window in ROH, and this state is inferred more confidently with more data. Overall, the model predicts ROH well in all cases, even at a coverage of 0.1x (Fig.\ref{fig2:ROH}).

@@

\paragraph{IBD prediction}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness and ranges from 1.0 to 0.563 at 0.03x. On the other hand, the accuracy for identical individuals and parent-child pairs is nearly perfect for all investigated coverages, even at 0.03x coverage.
[add some comments about contamination and ascertainment bias, with reference to supplementary figures]

@@

\paragraph{Relatedness classification}
We finally evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig. \ref{fig3:Comparison_READ_KIN}). We first describe the results for relatedness cases detectable by READ, viz. identical, $1_{st} degree$, $2_{nd} degree$ and $unrelated$. In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data. The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02.  One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.94 to 0.86, compared to range of 0.91 to 0.74 for READ). The false positive rate in this case remains below 0.02 for both methods. We also find that READ is strongly impacted by contamination ($<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.53 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively. By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control. For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives. The increases in false positives is likely due to ROH making individuals more similar (for details, see table S4). Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e. parent-child, siblings and, to a lower level, 3rd degree relatives (Fig. \ref{fig3:Comparison_READ_KIN})

@@

\subsubsection{Application to real data}

\paragraph{Chagyrskaya Neandertals}
To test KIN on real ancient data, we applied it to a Neandertal dataset from Chagyrskaya Cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020,laurits_skov_genetic_nodate}. This dataset contains genetic data from  14 skeletal remains that likely belong to contemporary Neandertals who occupied the cave between 63 ± 4 and 48 ± 3 kya \cite{laurits_skov_genetic_nodate}. DNA extracted from each of these remains were captured with an array targeting variable sites identified in high coverage Neandertal and Denisovan genomes, and common variations in Africans \cite{laurits_skov_genetic_nodate}. This genetic data has  low-to-intermediate depth of coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and DNA contamination from modern humans as well as hyenas \cite{laurits_skov_genetic_nodate}. We focused our analysis to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017} as done by the authors for the relatedness analysis \cite{laurits_skov_genetic_nodate}. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three specimens from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). Further, we identify Chagyrskaya17 and Chagyrskaya60 as 3rd-degree relatives.

@@

\subsubsection{Ancient modern humans}

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin. We found that in most cases the three methods agree. Only $<5\%$ of pairs (315 out of 6903) show a different relatedness for KIN and READ (table S2). However, most of these are at very low confidence for both the methods ($\Delta LL>1$ for KIN and $sd<1$ for READ). There are 28 pairs where KIN shows confident results, and when we plot the raw heterozygosity estimates along the genome, in many cases the prediction by KIN is in agreement with the observed proportion of differences along the genome (fig. S\ref{figS8:eg1}). Comparisons with lcMLkin are spare since lcMLkin shows reliable results for $\appr 53\%$ pairs, and we find 16 pairs for which lcMLkin gives a reliable result (\cite{mittnik_kinship-based_2019}), and KIN shows $\Delta LL>$1. We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq20,000$), we can clearly see that proportion of differences along the genome aligns with prediction of KIN (fig. S\ref{figS9:eg2}).
